Acceleration of plague outbreaks in the second pandemic\\\emph{Electronic \supp}

### 2020-07-11 supp.Rnw
ref to KeelGill00 for plague natural history
hyperlink

\paragraph{London 1593:} The extant records of weekly plague mortality in London in 1593 (\cref{F:plot3sourcesmajors}, second panel) are implausible.  One possibility is that many earlier deaths were added to the counts for two of the weeks in July, but there is no way to be sure.  Creighton \cite[pp.\,351--360]{Crei65} discusses all the data ever found for the epidemic of plague in 1592--1593, states that the weekly data probably originate from marginal notes in a broadside of 1603, and comments (p.\,354) that ``the weekly mortalities in it for those weeks that had little plague are an absurdity for 1593.  Whatever the source of this table, it is not genuine for 1593 \dots''

@@

\cref{F:all1600s.scaled} shows all weekly reports of plague deaths from the LBoM from 1563 to 1666, scaled by population size (estimated by interpolating from \cref{tab:pop}).  Epidemics with peak plague mortality above 5 per 1000 individuals per week were classified as \emph{major}.  With the exception of 1593 (for which we do not use LBoM data; see above), all these major epidemics peak above 14 on this scale; all the other (\emph{minor}) epidemics peak below 3.  The raw plague mortality data are shown in \cref{F:all1600s}.

@@

For pneumonic plague, sufficiently detailed data exist for a number of modern outbreaks to allow an estimate of the latent and infectious periods (and hence of the generation interval distribution) \cite{GaniLeac04}, which we can use to estimate $\rzero$ for a given value of $r$. Much less data is available for bubonic plague; in addition, the more complex host-vector life cycle of bubonic plague complicates the estimation of the generation interval. However, we can say that the elapsed time between the onset of infectiousness of a rat and the time when a rat in the next infection cycle becomes infectious is:
%%
\begin{equation}
  \begin{split}
    &\quad \textrm{rat} \to \textrm{flea infection time} \\
    \mbox{} + &\quad \textrm{max}\bigl( \textrm{flea incubation period},
    \{\textrm{time to rat death} + \textrm{flea searching time}\} \bigr) \\
    \mbox{} + &\quad \textrm{flea} \to \textrm{rat infection time} \\
    \mbox{} + &\quad \textrm{rat incubation period}
  \end{split}
\end{equation}
%%
The second line above takes account of the fact that a flea leaves
its rat host when, and only when, the rat dies.  From
\cite{LaPerriere+2009}, we can gather that
\begin{itemize}
\item fleas bite rats $\approx$ 4 times/day; the rat $\to$ flea infection probability is $\approx$ 0.2 and the flea $\to$ rat infection probability is $\approx$ 0.28, suggesting that both the rat $\to$ flea and flea $\to$ rat infection times are $\approx$ 1 day; 
\item the rat infectious period (time until infected rats die) is $\approx$ 4 days (\cite{LaPerriere+2009} cite \cite{Audi03} for this value; \cite{KeelGill00b} give a value of $\approx$ 18 days, which seems unrealistically long --- another source, \cite{Toll+10}, gives values ranging from $\approx$ 4--9 days)
\item the flea incubation period is long (but very variable), ranging from 9--26 days (\cite{LaPerriere+2009}, citing \cite{Poll52})
\item the flea searching time is not explicitly defined by \cite{LaPerriere+2009}, but we guess it is relatively short ($<$ 1--2 days)
\item \cite{LaPerriere+2009} give a value of $\approx$ 1--3 days for the rat incubation period
\end{itemize}
%%
Since the flea incubation period is typically longer than the combination of rat death time and flea searching time (9--26 days vs.\ (4--9 + 1--2) days), we can approximate the generation interval as $\approx$ (1 day) + (9--26 days) + (1 day) + (1--3 days); we use a value of $\gentime=18$ days in the main text.

>>>>

\hypertarget{1593}{}
\paragraph{London 1593:} The extant records of weekly plague mortality in London in 1593 (\cref{F:plot3sourcesmajors}, second panel) are implausible.  One possibility is that many earlier deaths were added to the counts for two of the weeks in July, but there is no way to be sure.  Creighton \cite[pp.\,351--360]{Crei65} discusses all the data ever found for the epidemic of plague in 1592--1593, states that the weekly data probably originate from marginal notes in a broadside of 1603, and comments (p.\,354) that ``the weekly mortalities in it for those weeks that had little plague are an absurdity for 1593.  Whatever the source of this table, it is not genuine for 1593 \dots''

@@

\cref{F:all1600s.scaled} shows all weekly reports of plague deaths from the LBoM from 1563 to 1666, scaled by population size (estimated by interpolating from \cref{tab:pop}).  Epidemics with peak plague mortality above 5 per 1000 individuals per week were classified as \emph{major}.  With the exception of 1593 (for which we do not use LBoM data; see \hyperlink{1593}{above}), all these major epidemics peak above 14 on this scale; all the other (\emph{minor}) epidemics peak below 3.  The raw plague mortality data are shown in \cref{F:all1600s}.

@@

For pneumonic plague, sufficiently detailed data exist for a number of modern outbreaks to allow an estimate of the latent and infectious periods (and hence of the generation interval distribution) \cite{GaniLeac04}, which we can use to estimate $\rzero$ for a given value of $r$. Much less data is available for bubonic plague; in addition, the more complex host-vector life cycle of bubonic plague complicates the estimation of the generation interval. However, we can say that the elapsed time between the onset of infectiousness of a rat and the time when a rat in the next infection cycle becomes infectious is:
%%
\begin{equation}
  \begin{split}
    &\quad \textrm{rat} \to \textrm{flea infection time} \\
    \mbox{} + &\quad \textrm{max}\bigl( \textrm{flea incubation period},
    \{\textrm{time to rat death} + \textrm{flea searching time}\} \bigr) \\
    \mbox{} + &\quad \textrm{flea} \to \textrm{rat infection time} \\
    \mbox{} + &\quad \textrm{rat incubation period}
  \end{split}
\end{equation}
%%
The second line above takes account of the fact that a flea leaves
its rat host when, and only when, the rat dies \cite{KeelGill00}.  From
\cite{LaPerriere+2009}, we can gather that
\begin{itemize}
\item fleas bite rats $\approx$ 4 times/day; the rat $\to$ flea infection probability is $\approx$ 0.2 and the flea $\to$ rat infection probability is $\approx$ 0.28, suggesting that both the rat $\to$ flea and flea $\to$ rat infection times are $\approx$ 1 day; 
\item the rat infectious period (time until infected rats die) is $\approx$ 4 days (\cite{LaPerriere+2009} cite \cite{Audi03} for this value; \cite{KeelGill00b} give a value of $\approx$ 18 days, which seems unrealistically long --- another source, \cite{Toll+10}, gives values ranging from $\approx$ 4--9 days)
\item the flea incubation period is long (but very variable), ranging from 9--26 days (\cite{LaPerriere+2009}, citing \cite{Poll52})
\item the flea searching time is not explicitly defined by \cite{LaPerriere+2009}, but we guess it is relatively short ($<$ 1--2 days)
\item \cite{LaPerriere+2009} give a value of $\approx$ 1--3 days for the rat incubation period
\end{itemize}
%%
Since the flea incubation period is typically longer than the combination of rat death time and flea searching time (9--26 days vs.\ (4--9 + 1--2) days), we can approximate the generation interval as $\approx$ (1 day) + (9--26 days) + (1 day) + (1--3 days); we use a value of $\gentime=18$ days in the main text.
